Implementation and comparison of register allocation to translate \lstinline!LLVM--! to x86

### 2023-12-20 article.tex
Jane approved

The simplest ones to implement are the \lstinline!def! and \lstinline!use! functions, as all of the values of interest are located immediately within the instruction itself and not hidden behind some layer of indirection:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let def (s : S.SS.t) (insn : Cfg.insn) =
  match insn with Insn (Some dop, _) -> S.SS.add dop s | _ -> s

@@

The purpose for conducting dataflow analysis as above is finding variables that may be assigned the same register. This is done by building an interference graph, which is an undirected graph, whose nodes represent variables and edges signify interference between them, i.e. variables \(a\) and \(b\) live at overlapping program points is represented with an edge \((a,b)\).

@@

Because of this, for any phi node of the form 
\[
  a = \Phi (b_1, ..., b_n)
\]
add edges to all \textit{live-out} variables not in \(B\)
\[
  \forall b_j \in (\mathit{out}[i] \setminus B), \mathit{add\_edge}(a, b_j) \text{ where } B = \{b_1, ..., b_n\}
\]
For any other instruction the defines a variable \(a\)
\[
  \forall b_j \in B, \mathit{add\_edge}(a, b_j)
\]
Although the Appel text notes interference with concrete registers as well as overlapping variables, this isn't considered in this implementation.

@@

%* list of methods for coalescing/allocating:
%   * ocamlgraph builtin
%   * greedy 
%   * briggs
%   * george
%   * welsh-powell?
\section{Graph Coloring}
Once an interference graph is constructed, the actual assignments can be found using graph coloring. Although this has long been known to be NP-complete, the heuristic as introduced in both the Appel and Aho et al. texts is a linear time approximation to this problem. It is based on an iterative approach wherein nodes known to be colorable are removed until either an empty graph remains in which case the original graph \(G\) is \(k\)-colorable  or nodes with more than \(k\) neighbours remain, in which case a node is chosen to be spilled to the stack and removed. This is then repeated.

Let \(k\) be the number of working registers available on the target architecture. After building the interference graph \(G\), a node \(n\) with fewer than \(k\) neighbors is chosen. As \(n\) has at most \(k-1\) neighbors, it can be removed safely, effectively simplifying \(G\). It is pushed to a stack in order to preserve the order in which they are removed so that \(G\) can be rebuilt and the corresponding registers can be assigned correctly once a \(k\)-colorable assignment is found.
%  (i.e. \(\mathit{deg}(n) < k\))

@@

Coalescing is the process of eliminating moves/copies of data from one GPR to another by combining their interference graph nodes. This is similar to but  not the same as  the lack of interference edges between variables subject to move operations. This is because variables \(a\) and \(b\) may still be assigned different registers or even spilled if,  for instance, either of them are of significant degree. Coalescing joins nodes \(a\) and \(b\) to node \(ab\) preserving the edges of both to maintain soundness.

Since all edges are preserved, the resulting node \(an\) may be of a much higher degree. Because to this, only strategies that produce a \(k\)-colorable graph are worth considering as worth considering as additional spills negate the purpose entirely.

@@

@@

Validating the correctness of each instruction translation is a complicated task given how low-level it is. One approach is attaching a debugger and examining the state before and after the translated sequence of x86 instructions is executed. The result of applying a binary operation should only affect the assigned register, for instance.

>>>>

The simplest equations to implement are the \lstinline!def! and \lstinline!use! functions, as all of the values of interest are located immediately within the instruction itself and not hidden behind some layer of indirection:
\begin{minted}[linenos,bgcolor=LightGray]{ocaml}
let def (s : S.SS.t) (insn : Cfg.insn) =
  match insn with Insn (Some dop, _) -> S.SS.add dop s | _ -> s

@@

The purpose for conducting dataflow analysis as above is finding variables that may be assigned the same register. This is done by building an interference graph, which is an undirected graph, whose nodes represent variables and edges that signify interference between them, i.e. variables \(a\) and \(b\) live at overlapping program points is represented with an edge \((a,b)\).

@@

Because of this, for any phi node of the form 
\[
  a = \Phi (b_1, ..., b_n)
\]
add edges to all \textit{live-out} variables not in \(B\):
\[
  \forall b_j \in (\mathit{out}[i] \setminus B), \mathit{add\_edge}(a, b_j) \text{ where } B = \{b_1, ..., b_n\}
\]
For any other instruction that defines a variable \(a\)
\[
  \forall b_j \in B, \mathit{add\_edge}(a, b_j).
\]
Although the Appel text (\cite{tiger}) describes interference of variables with concrete registers as well as overlapping variables, this isn't considered in this implementation.

@@

%* list of methods for coalescing/allocating:
%   * ocamlgraph builtin
%   * greedy 
%   * briggs
%   * george
%   * welsh-powell?
\section{Graph Coloring}
Once an interference graph is constructed, the actual assignments can be found using graph coloring. Although this has long been known to be NP-complete, the heuristic as introduced in both the Appel and Aho et al. texts is a linear time approximation to this problem. It is based on an iterative approach wherein nodes known to be colorable are removed until either an empty graph remains, in which case the original graph \(G\) is \(k\)-colorable  or nodes with more than \(k\) neighbours remain. In this case a node is chosen to be spilled to the stack and removed. This is then repeated.

Let \(k\) be the number of working registers available on the target architecture. After building the interference graph \(G\), a node \(n\) with fewer than \(k\) neighbors is chosen. As \(n\) has at most \(k-1\) neighbors, it can be removed safely, effectively simplifying \(G\). It is pushed to a stack to preserve the order in which they are removed so that \(G\) can be rebuilt and the corresponding registers can be assigned correctly once a \(k\)-colorable assignment is found.

%  (i.e. \(\mathit{deg}(n) < k\))

@@

Coalescing is the process of eliminating moves/copies of data from one GPR to another by combining their interference graph nodes.  This is similar to but  not the same as  the lack of interference edges between variables that are subject to move operations. This is because variables \(a\) and \(b\) may still be assigned different registers or even spilled if,  for instance, either of them are of significant degree. Coalescing joins nodes \(a\) and \(b\) to node \(ab\) preserving the edges of both to maintain soundness.

Since all edges are preserved, the resulting node \(ab\) may be of a much higher degree. Because of this, only strategies that produce a \(k\)-colorable graph are worth considering, as additional spills negate the purpose entirely.

@@

% TODO: find ud af om man skal henvise til kurser på au på speciel form

@@

Validating the correctness of each instruction translation is a complicated task given how low-level it is. One approach is attaching a debugger and examining the state before and after the translated sequence of x86 instructions is executed. The result of applying a binary operation,  for instance, should only affect the assigned register.
